---
title: "Bellabeat Case Study"
author: "oxbbar"
output:
  pdf_document: default
  html_document: default
---

## Setup

Install and load the required packages.
``` {r results='hide', message=FALSE}
install.packages("dplyr",repos = "http://cran.us.r-project.org")
install.packages("tidyr",repos = "http://cran.us.r-project.org")
install.packages("ggplot2",repos = "http://cran.us.r-project.org")
install.packages("lubridate",repos = "http://cran.us.r-project.org")
install.packages("readr",repos = "http://cran.us.r-project.org")
install.packages("janitor",repos = "http://cran.us.r-project.org")
install.packages("styler",repos = "http://cran.us.r-project.org")
install.packages("plotly",repos = "http://cran.us.r-project.org")
library(dplyr)
library(tidyr)
library(ggplot2)
library(lubridate)
library(readr)
library(janitor)
library(styler)
library(plotly)
```

Create a list of csv files in the directory.
``` {r}
files_csv <- list.files(path = "bellabeat_data/", pattern = "*.csv")
```

Create a list of file names for the data frames.
``` {r}
files <- substr(files_csv,1,nchar(files_csv)-4)
```

Create data frames for all the csv files.
``` {r results='hide', message=FALSE}
for(i in files){
  filepath = file.path(paste("bellabeat_data/",i,".csv", sep = ""))
  assign(i,read_csv(filepath))
}
```

Count the distinct IDs in each data frame to determine how many individuals there were for each data set.
``` {r}
n_distinct(dailyActivity_merged$Id)
n_distinct(dailyCalories_merged$Id)
n_distinct(dailyIntensities_merged$Id)
n_distinct(dailySteps_merged$Id)
n_distinct(heartrate_seconds_merged$Id)
n_distinct(hourlyCalories_merged$Id)
n_distinct(hourlyIntensities_merged$Id)
n_distinct(hourlySteps_merged$Id)
n_distinct(minuteCaloriesNarrow_merged$Id)
n_distinct(minuteCaloriesWide_merged$Id)
n_distinct(minuteIntensitiesNarrow_merged$Id)
n_distinct(minuteIntensitiesWide_merged$Id)
n_distinct(minuteMETsNarrow_merged$Id)
n_distinct(minuteSleep_merged$Id)
n_distinct(minuteStepsNarrow_merged$Id)
n_distinct(minuteStepsWide_merged$Id)
n_distinct(sleepDay_merged$Id)
n_distinct(weightLogInfo_merged$Id)
```

All data sets had 33 individuals, except for:
- heartrate_seconds = 14
- minuteSleep = 24
- sleepDay = 24
- WeightLogInfo = 8

### Processing the Daily Activity Data

Determine how the data is structured.
``` {r}
str(dailyActivity_merged)
```

Check for NA values and duplicate rows.
``` {r}
sum(is.na(dailyActivity_merged))
sum(duplicated(dailyActivity_merged))
```

Clean the column names.
``` {r}
dailyActivity_merged <- clean_names(dailyActivity_merged)
glimpse(dailyActivity_merged)
```

Add a new date column in appropriate format, column for day of the week (with Monday being designated as the first day) and an "id_and_date" to use as reference for joins later.
``` {r}
dailyActivity_merged <- dailyActivity_merged %>%
  mutate(
    new_date = as.Date(activity_date,"%m/%d/%y"),
    week_day = wday(new_date,label=TRUE,week_start=1),
    id_and_date = paste0(id,"_and_",new_date)
  )
str(dailyActivity_merged)
```

### Processing the Daily Sleep Data

Determine how the data is structured.
``` {r}
str(sleepDay_merged)
```

Check for NA values and duplicate rows.
``` {r}
sum(is.na(sleepDay_merged))
sum(duplicated(sleepDay_merged))
```

Remove duplicates.
``` {r}
sleepDay_merged <- sleepDay_merged[!duplicated(sleepDay_merged),]
sum(duplicated(sleepDay_merged))
```

Clean the column names.
``` {r}
sleepDay_merged <- clean_names(sleepDay_merged)
str(sleepDay_merged)
```

Add a new column for time awake in bed, a new date column in appropriate format, column for day of the week (with Monday being designated as the first day) and an "id_and_date" to use as reference for joins later.
``` {r}
sleepDay_merged <- sleepDay_merged %>%
  mutate(
    time_awake_in_bed = total_time_in_bed - total_minutes_asleep,
    new_date = as.Date(sleep_day,"%m/%d/%y"),
    week_day = wday(new_date,label=TRUE,week_start=1),
    id_and_date = paste0(id,"_and_",new_date)
  )
str(sleepDay_merged)
```

### Processing the Daily Weight Data

Determine how the data is structured.
``` {r}
str(weightLogInfo_merged)
```

Check for NA values and duplicate rows.
``` {r}
sum(is.na(weightLogInfo_merged))
sum(duplicated(weightLogInfo_merged))
```

NA values exist - double check what they actually are.
``` {r}
head(weightLogInfo_merged)
```
They are just 'Fat' records that weren't put in, but other daily data exists for these dates and, if we need, we can filter out these NA vales if we analyse Fat data.

Clean the column names.
``` {r}
weightLogInfo_merged <- clean_names(weightLogInfo_merged)
str(weightLogInfo_merged)
```

Add a new date column in appropriate format, column for day of the week and an "id_and_date" to use as reference for joins later.
``` {r}
weightLogInfo_merged <- weightLogInfo_merged %>%
  mutate(
    new_date = as.Date(date,"%m/%d/%y"),
    week_day = wday(new_date,label=TRUE),
    id_and_date = paste0(id,"_and_",new_date)
  )
str(weightLogInfo_merged)
```

### Creating temp tables for joining

Daily Activity Data.
``` {r}
colnames(dailyActivity_merged)
```
``` {r}
temp_daily_activity <- select(dailyActivity_merged, 
  "id", 
  "new_date",
  "id_and_date",
  "total_steps", 
  "total_distance", 
  "very_active_distance",
  "moderately_active_distance",
  "light_active_distance",
  "sedentary_active_distance",
  "very_active_minutes",
  "fairly_active_minutes",
  "lightly_active_minutes",
  "sedentary_minutes",
  "calories", 
  "week_day"
  )
```

Daily Sleep Data.
``` {r}
colnames(sleepDay_merged)
```
``` {r}
temp_daily_sleep <- select(sleepDay_merged, 
  "id_and_date",
  "total_sleep_records",
  "time_awake_in_bed",
  "total_minutes_asleep",
  "total_time_in_bed",
  )
```

Merge these two data frames with an outer join.
``` {r}
merged_daily_activity_and_sleep <- merge(x = temp_daily_activity, y = temp_daily_sleep, by = "id_and_date", all = TRUE)
str(merged_daily_activity_and_sleep)
```

Daily Weight Data.
``` {r}
colnames(weightLogInfo_merged)
```
``` {r}
temp_daily_weight <- select(weightLogInfo_merged, 
  "id_and_date",
  "weight_kg",
  "weight_pounds",
  "fat",
  "bmi",
  "is_manual_report",
  )
colnames(temp_daily_weight)
```

Merge all data together.
``` {r}
merged_data_daily <- merge(x = merged_daily_activity_and_sleep, y = temp_daily_weight, by = "id_and_date", all = TRUE)
str(merged_data_daily)
glimpse(merged_data_daily)
```

### Analysing the Daily Data

Create a data frame of averages to analyse step data and intensity of distances and minutes.
``` {r}
merged_data_daily_averages <- merged_data_daily %>%
  ## remove records where no steps were taken (no device usage)
  filter(total_steps > 0) %>%
  group_by(week_day) %>%
    summarise(
      mean_steps = mean(total_steps),
      mean_total_distance = mean(total_distance),
      mean_very_active_distance = mean(very_active_distance),
      mean_moderately_active_distance = mean(moderately_active_distance),
      mean_light_active_distance = mean(light_active_distance),
      mean_sedentary_active_distance = mean(sedentary_active_distance),
      mean_very_active_minutes = mean(very_active_minutes),
      mean_fairly_active_minutes = mean(fairly_active_minutes),
      mean_lightly_active_minutes = mean(lightly_active_minutes),
      mean_sedentary_minutes = mean(sedentary_minutes),
      mean_calories = mean(calories),
      total_records = sum(!is.na(week_day))
    )
glimpse(merged_data_daily_averages)
```

Create a graph of total steps for all users per week day.
``` {r}
all_steps_daily <- merged_data_daily %>%
  group_by(week_day) %>%
  summarise(
    all_steps = sum(total_steps)
  )
ggplot(data=all_steps_daily,aes(x=week_day,y=all_steps,fill=all_steps)) + geom_col() + geom_col() + labs(title="Total Steps Taken for All Users per Week Day", x="Week Day", y="Total Steps") + guides(fill="none")
```

Create a graph of average steps per week day.
``` {r}
ggplot(data=merged_data_daily_averages, aes(x=week_day,y=mean_steps,fill=mean_steps)) + geom_col() + labs(title="Average Steps Taken Per Week Day", x="Week Day", y="Average Steps") + guides(fill="none")
```

Create a graph of average calories per week day.
``` {r}
ggplot(data=merged_data_daily_averages, aes(x=week_day,y=mean_calories,fill=mean_calories)) + geom_col() + labs(title="Average Calories Burned Per Week Day", x="Week Day", y="Average Calories") + guides(fill="none")
```

Create a graph of average records per week day.
``` {r}
ggplot(data=merged_data_daily_averages, aes(x=week_day,y=total_records,fill=total_records)) + geom_col() + labs(title="Average Data Entries Per Week Day", x="Week Day", y="Average Data Entries") + guides(fill="none")
```

Create a graph of average sedentary minutes per week day.
``` {r}
average_sedentary_daily <- merged_data_daily %>%
filter(total_steps > 0) %>%
  group_by(week_day) %>%
  summarise(
    mean_sed_mins = mean(sedentary_minutes)
  )

ggplot(data=average_sedentary_daily, aes(x=week_day,y=mean_sed_mins,fill=mean_sed_mins)) + geom_col() + labs(title="Average Sedentary Minutes Per Week Day", x="Week Day", y="Average Sedentary Minutes") + guides(fill="none")
```

Create a graph of average lightly active minutes per week day.
``` {r}
average_lightly_active_daily <- merged_data_daily %>%
filter(total_steps > 0) %>%
  group_by(week_day) %>%
  summarise(
    mean_la_mins = mean(lightly_active_minutes)
  )

ggplot(data=average_lightly_active_daily, aes(x=week_day,y=mean_la_mins,fill=mean_la_mins)) + geom_col() + labs(title="Average Lightly Active Minutes Per Week Day", x="Week Day", y="Average Lightly Active Minutes") + guides(fill="none")
```

Create a graph of average fairly active minutes per week day.
``` {r}
average_fairly_active_daily <- merged_data_daily %>%
filter(total_steps > 0) %>%
  group_by(week_day) %>%
  summarise(
    mean_fa_mins = mean(fairly_active_minutes)
  )

ggplot(data=average_fairly_active_daily, aes(x=week_day,y=mean_fa_mins,fill=mean_fa_mins)) + geom_col() + labs(title="Average Fairly Active Minutes Per Week Day", x="Week Day", y="Average Fairly Active Minutes") + guides(fill="none")
```

Create a graph of average very active minutes per week day.
``` {r}
average_very_active_daily <- merged_data_daily %>%
filter(total_steps > 0) %>%
  group_by(week_day) %>%
  summarise(
    mean_va_mins = mean(very_active_minutes)
  )

ggplot(data=average_very_active_daily, aes(x=week_day,y=mean_va_mins,fill=mean_va_mins)) + geom_col() + labs(title="Average Very Active Minutes Per Week Day", x="Week Day", y="Average Very Active Minutes") + guides(fill="none")
```

Create a graph of daily steps vs. daily calories burned.
``` {r}
daily_steps_vs_calories <- merged_data_daily %>%
  summarise (
    total_steps = total_steps,
    calories = calories
  )
ggplot(data=daily_steps_vs_calories, aes(x=total_steps,y=calories)) + geom_smooth(method = lm, aes(x=total_steps,y=calories)) + geom_point() + labs(title="Total Steps taken vs. Calories Burned Per Day", x="Total Steps", y="Calories Burned")
```

Create a graph of very active minutes vs. daily calories burned.
``` {r}
very_active_calories <- merged_data_daily %>%
summarise (
    very_active_minutes = very_active_minutes,
    calories = calories    
    )
ggplot(data=very_active_calories, aes(x=very_active_minutes,y=calories)) + geom_point() + geom_smooth(method=lm, aes(x=very_active_minutes,y=calories)) + labs(title="Very Active Minutes vs. Calories Burned", x="Very Active Minutes",y="Calories Burned")
```

Create a graph of sedentary minutes vs. daily calories burned.
``` {r}
sedentary_calories <- merged_data_daily %>%
summarise (
    sedentary_minutes = sedentary_minutes,
    calories = calories    
    )
ggplot(data=sedentary_calories, aes(x=sedentary_minutes,y=calories)) + geom_point() + geom_smooth(method=lm, aes(x=sedentary_minutes,y=calories)) + labs(title="Sedentary Minutes vs. Calories Burned", x="Sedentary Minutes",y="Calories Burned")
```

Create a pie chart of percentage per day in each active zone.
``` {r}
active_minutes_summary <- merged_data_daily %>%
  summarise(
  sedentary_minutes = mean(sedentary_minutes),
  lightly_active_minutes = mean(lightly_active_minutes),
  fairly_active_minutes = mean(fairly_active_minutes),
  very_active_minutes = mean(very_active_minutes),
)

percentage <- data.frame(
  levels = c(
    "Sedentary",
    "Lightly",
    "Fairly Active",
    "Very Active"),
  percentage = c(
    active_minutes_summary$sedentary_minutes, 
    active_minutes_summary$lightly_active_minutes, 
    active_minutes_summary$fairly_active_minutes, 
    active_minutes_summary$very_active_minutes)
)

plot_ly(percentage, labels = ~levels, values = ~percentage, type = 'pie') %>%
  layout(title="Percentage of Day in Active Zones")
```

Create a data frame of averages to analyse sleep data, including time in bed not asleep.
``` {r}
merged_data_daily_sleep_averages <- merged_data_daily %>%
  ## remove records where no sleep record was made
  filter(total_sleep_records > 0) %>%
  group_by(week_day) %>%
    summarise(
      mean_time_awake_in_bed = mean(time_awake_in_bed),
      mean_total_minutes_asleep = mean(total_minutes_asleep),
      mean_total_time_in_bed = mean(total_time_in_bed),
      total_records = sum(!is.na(week_day))
    )
glimpse(merged_data_daily_sleep_averages)
```

Create a graph of average minutes asleep each week day.
``` {r}
merged_data_daily_sleep_averages_labelled <- merged_data_daily %>%
  ## remove records where no sleep record was made (no device usage)
  filter(total_sleep_records > 0) %>%
  group_by(week_day) %>%
  summarise(
    mean_total_minutes_asleep = mean(total_minutes_asleep),
    hours = (mean_total_minutes_asleep/60),
    floorhours = floor(hours),
    sumhours = round(hours %% floorhours * 60,0 ),
    label = paste0(floorhours,":",sumhours)
  )
ggplot(data=merged_data_daily_sleep_averages_labelled, aes(x=week_day,y=hours,label = label)) + geom_col() + geom_label(position = position_dodge(width = 1)) + xlab(label = "Week Day") + ylab(label = "Hours Sleep") + ggtitle(label = "Average Hours Sleep Each Week Day")
```


Create a graph of average time in bed and awake each week day.
``` {r}
merged_data_daily_bed_awake_averages_labelled <- merged_data_daily %>%
  ## remove records where no sleep record was made (no device usage)
  filter(total_sleep_records > 0) %>%
  group_by(week_day) %>%
  summarise(
    mean_time_awake_in_bed = mean(time_awake_in_bed),
    label = paste0(round(mean_time_awake_in_bed)," mins")
  )
ggplot(data=merged_data_daily_bed_awake_averages_labelled, aes(x=week_day,y=mean_time_awake_in_bed,label = label)) + geom_col() + geom_label(position = position_dodge(width = 1)) + xlab(label = "Week Day") + ylab(label = "Mins Awake in Bed") + ggtitle(label = "Average Minutes Awake in Bed Per Week Day")
```

### Processing the Hourly Steps Data

Determine how the data is structured.
``` {r}
str(hourlySteps_merged)
```

Check for NA values and duplicate rows.
``` {r}
sum(is.na(hourlySteps_merged))
sum(duplicated(hourlySteps_merged))
```

Clean the column names.
``` {r}
hourlySteps_merged <- clean_names(hourlySteps_merged)
str(hourlySteps_merged)
```

Add columns for the date, the hour(and ways to represent it) and id/date/hour for joins later.
``` {r}
hourlySteps_merged <- hourlySteps_merged %>%
  mutate(
    new_date = as.Date(activity_hour,"%m/%d/%y"),
    new_hour = format(mdy_hms(activity_hour),format="%H:%M"),
    hour_int = hour(mdy_hms(activity_hour)),
    hour_label = format(mdy_hms(activity_hour),format="%H%p"),
    hour_label_simple = format(mdy_hms(activity_hour),format="%H"),
    week_day = wday(new_date,label=TRUE,week_start=1),
    id_and_time = paste0(id,"_and_",new_date,"_and_",new_hour)
  )
str(hourlySteps_merged)
```

### Processing the Hourly Intensities Data

Determine how the data is structured.
``` {r}
str(hourlyIntensities_merged)
```

Check for NA values and duplicate rows.
``` {r}
sum(is.na(hourlyIntensities_merged))
sum(duplicated(hourlyIntensities_merged))
```

Clean the column names.
``` {r}
hourlyIntensities_merged <- clean_names(hourlyIntensities_merged)
str(hourlyIntensities_merged)
```

Add columns for the date, the hour and id/date/hour for joins later.
``` {r}
hourlyIntensities_merged <- hourlyIntensities_merged %>%
  mutate(
    new_date = as.Date(activity_hour,"%m/%d/%y"),
    new_hour = format(mdy_hms(activity_hour),format="%H:%M"),
    week_day = wday(new_date,label=TRUE,week_start=1),
    id_and_time = paste0(id,"_and_",new_date,"_and_",new_hour)
  )
str(hourlyIntensities_merged)
```

### Processing the Hourly Calories Data

Determine how the data is structured.
``` {r}
str(hourlyCalories_merged)
```

Check for NA values and duplicate rows.
``` {r}
sum(is.na(hourlyCalories_merged))
sum(duplicated(hourlyCalories_merged))
```

Clean the column names.
``` {r}
hourlyCalories_merged <- clean_names(hourlyCalories_merged)
str(hourlyCalories_merged)
```

Add columns for the date, the hour and id/date/hour for joins later.
``` {r}
hourlyCalories_merged <- hourlyCalories_merged %>%
  mutate(
    new_date = as.Date(activity_hour,"%m/%d/%y"),
    new_hour = format(mdy_hms(activity_hour),format="%H:%M"),
    week_day = wday(new_date,label=TRUE),
    id_and_time = paste0(id,"_and_",new_date,"_and_",new_hour)
  )
str(hourlyCalories_merged)
```

### Creating temp tables for joining

Hourly Steps Data.
``` {r}
colnames(hourlySteps_merged)
```
``` {r}
temp_hourly_steps <- select(hourlySteps_merged, 
  "id", 
  "activity_hour",
  "id_and_time",
  "new_date",
  "new_hour",
  "hour_int",
  "hour_label",
  "hour_label_simple",
  "week_day",
  "step_total"
  )
```

Hourly Intensities Data.
``` {r}
colnames(hourlyIntensities_merged)
```
``` {r}
temp_hourly_intensities <- select(hourlyIntensities_merged, 
  "id_and_time",
  "total_intensity",
  "average_intensity"
  )
```

Merge these two data frames with an outer join.
``` {r}
merged_hourly_steps_and_intensities <- merge(x = temp_hourly_steps, y = temp_hourly_intensities, by = "id_and_time", all = TRUE)
str(merged_hourly_steps_and_intensities)
```

Hourly Calories Data.
``` {r}
colnames(hourlyCalories_merged)
```
``` {r}
temp_hourly_calories <- select(hourlyCalories_merged, 
  "id_and_time",
  "calories"
  )
colnames(temp_hourly_calories)
```

Merge all data together.
``` {r}
merged_data_hourly <- merge(x = merged_hourly_steps_and_intensities, y = temp_hourly_calories, by = "id_and_time", all = TRUE)
str(merged_data_hourly)
glimpse(merged_data_hourly)
```

### Analysing the Hourly Data

Create a graph of average steps per hour for all days.
``` {r}
merged_data_hourly_averages <- merged_data_hourly %>%
  group_by(hour_label_simple) %>%
  summarise (
    mean_step_total = mean(step_total),
  ) 
glimpse(merged_data_hourly_averages)
ggplot(data=merged_data_hourly_averages, aes(x=hour_label_simple, y=mean_step_total, fill=mean_step_total)) + geom_col() + xlab(label = "Hour") + ylab(label = "Average Steps") + ggtitle(label = "Average Steps per Hour for All Days") + guides(fill="none")
```

Create graphs of average steps per hour for each week day.
``` {r}
merged_data_hourly_averages <- merged_data_hourly %>%
  group_by(week_day, hour_label_simple) %>%
  summarise (
    mean_step_total = mean(step_total),
  ) 
glimpse(merged_data_hourly_averages)
ggplot(data=merged_data_hourly_averages, aes(x=hour_label_simple, y=mean_step_total, fill=hour_label_simple)) + geom_col() + xlab(label = "Hour") + ylab(label = "Average Steps") + ggtitle(label = "Average Steps per Hour for Each Days") + facet_wrap(~week_day) + guides(fill="none")
```

Create a graph of average calories burned per hour for all days.
``` {r}
hourly_calories <- merged_data_hourly %>%
  group_by(hour_label_simple) %>%
  summarise (
    mean_calories = mean(calories),
  ) 

glimpse(hourly_calories)

ggplot(data=hourly_calories, aes(x=hour_label_simple, y=mean_calories,fill=mean_calories)) + geom_col() + xlab(label = "Hour") + ylab(label = "Average Steps") + ggtitle(label = "Average Calories per Hour for All Days")
```

Create graphs of average calories burned per hour for each week day.
``` {r}
hourly_calories <- merged_data_hourly %>%
  group_by(week_day, hour_label_simple) %>%
  summarise (
    mean_calories = mean(calories),
  ) 

glimpse(hourly_calories)

ggplot(data=hourly_calories, aes(x=hour_label_simple, y=mean_calories,fill=mean_calories)) + geom_col() + facet_wrap(~week_day) + xlab(label = "Hour") + ylab(label = "Average Steps") + ggtitle(label = "Average Calories per Hour for All Days")
```

Create a graph of average intensity per hour vs. calories burned per hour.
``` {r}
average_vs_cal <- merged_data_hourly %>%
  summarise (
    average_intensity = average_intensity,
      calories = calories
  ) 

ggplot(data=average_vs_cal, aes(x=average_intensity, y=calories)) + geom_smooth(aes(x=average_intensity, y=calories)) + geom_point() + xlab(label = "Average Intensity") + ylab(label = "Calories Burned") + ggtitle(label = "Average Intensities Per Hour vs. Calories Burned")
```